Stop Lying to Me: New Visual Tools to Choose the Most Honest Nonlinear Dimension Reduction

### 2025-06-26 appendix.qmd
PBMC example revised - PH

Note that @xia2023 uses a different subset from the PBMC dataset used by @chen2024, shown by us in the main paper example, which is why this comparison is shown here and not in the main paper. Their data contains $31,021$ cells including cell type labels, and the gene expression levels were in the unit of log-transformed UMI count per $10,000$. They focused on three sequencing methods (inDrops, DropSeq, and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell, CD14+ Monocyte, and B cell. Pre-processing follows the process in @xia2023 again using the Human Peripheral Blood Mononuclear Cells (PBMC) data.

>>>>

Note that @xia2023 uses a different PBMC dataset than that used by @chen2024, shown by us in the main paper example, which is why this comparison is shown here and not in the main paper. Their data contains $31,021$ cells including cell type labels, and the gene expression levels were in the unit of log-transformed UMI count per $10,000$. They focused on three sequencing methods (inDrops, DropSeq, and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell, CD14+ Monocyte, and B cell. Pre-processing follows the process in @xia2023 again using the Human Peripheral Blood Mononuclear Cells (PBMC) data.

### 2025-06-26 appendix.tex
Added True model projections to appendix
PBMC example revised - PH

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$}
\newcommand\kD{$k\text{-}D$}
\newcommand\dD{$d\text{-}D$}
\newcommand\gD{$2\text{-}D$}
\newcommand{\To}{\textbf{to}~}
\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Appendix: Looking at Non-Linear Dimension Reductions as Models in the Data Space},
  pdfauthor={Jayani P. Gamage; Dianne Cook; Paul Harrison; Michael Lydeamore; Thiyanga S. Talagala},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\section{Generating the nonlinear clusters
data}\label{generating-the-nonlinear-clusters-data}

@@

Note that \citet{xia2023} uses a different subset from the PBMC dataset
used by \citet{chen2024}, shown by us in the main paper example, which
is why this comparison is shown here and not in the main paper. Their
data contains \(31,021\) cells including cell type labels, and the gene
expression levels were in the unit of log-transformed UMI count per
\(10,000\). They focused on three sequencing methods (inDrops, DropSeq,
and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell,
CD14+ Monocyte, and B cell. Pre-processing follows the process in
\citet{xia2023} again using the Human Peripheral Blood Mononuclear Cells
(PBMC) data.

>>>>

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$}
\newcommand\kD{$k\text{-}D$}
\newcommand\dD{$d\text{-}D$}
\newcommand\gD{$2\text{-}D$}
\newcommand\tD{$3\text{-}D$}
\newcommand\fD{$4\text{-}D$}
\newcommand\sD{$7\text{-}D$}
\newcommand{\To}{\textbf{to}~}
\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Appendix: Looking at Non-Linear Dimension Reductions as Models in the Data Space},
  pdfauthor={Jayani P. Gamage; Dianne Cook; Paul Harrison; Michael Lydeamore; Thiyanga S. Talagala},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\section{Generating the 2NC7 data}\label{generating-the-2nc7-data}

This data is constructed by simulating two clusters, each consisting of
\(1000\) observations. The C-shaped cluster is generated from
\(\theta \sim U(\text{-}3\pi/2, 0)\), \(X_1 = \sin(\theta)\),
\(X_2 \sim U(0, 2)\) (adding thickness to the C),
\(X_3 = \text{sign}(\theta) \times (\cos(\theta) - 1)\),
\(X_4 = \cos(\theta)\). Observations lie on a \gD{} manifold in \sD{}.
The other cluster is from \(X_1 \sim U(0, 2)\), \(X_2 \sim U(0, 3)\),
\(X_3 = \text{-}(X_1^3 + X_2)\), and \(X_4 \sim U(0, 2)\). It is also
curved, but observations lie on a \tD{} manifold in\sD{}. Three more
variables, \(X_5, X_6, X_7\), that are small amounts of pure noise are
added. We would consider \(T=(X_1, X_2, X_3, X_4)\) to be the geometric
structure (true model) that we hope to capture.

{\centering \includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{appendix_files/figure-pdf/unnamed-chunk-4-1.pdf}

\caption{Two projections of the \pD{} true model overlaying the data are
shown in a, b.}

@@

Note that \citet{xia2023} uses a different PBMC dataset than that used
by \citet{chen2024}, shown by us in the main paper example, which is why
this comparison is shown here and not in the main paper. Their data
contains \(31,021\) cells including cell type labels, and the gene
expression levels were in the unit of log-transformed UMI count per
\(10,000\). They focused on three sequencing methods (inDrops, DropSeq,
and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell,
CD14+ Monocyte, and B cell. Pre-processing follows the process in
\citet{xia2023} again using the Human Peripheral Blood Mononuclear Cells
(PBMC) data.

### 2025-06-26 paper.qmd
PBMC example revised - PH

In this data there are $2622$ single cells and $1000$ gene expressions (variables). Following their pre-processing, different NLDR techniques were performed on the first nine principal components. @fig-NLDR-variety shows this data using a variety of methods, and different hyper-parameters. You can see that the result is wildly different depending on the choices. Layout a is a reproduction of the layout published in @chen2024. This layout suggests that the data has three very well separated clusters, each with an odd shape. The question is whether this accurately represents the cluster structure in the data, or whether they should have chosen b or c or d or e or f or g or h. This is what our new method can help with -- to decide which is the more accurate \gD{} representation of the cluster structure in the \pD{} data.

@@

To further assess the choices, we need to look at the model in the data space, by using a tour to show the wireframe model overlaid on the data in the $9\text{-}D$ space (@fig-model-pbmc-author-proj). Here we compare the published layout (a) versus what we argue is the best layout (e). The top row (a1, a2, a3) correspond to the published layout and the bottom row (e1, e2, e3) correspond to the optimal choice according to our procedure. The middle and right plots show two projections. The primary difference between the two models is that, the model of layout e does not fill out to the extent of the data but concentrates in the center of each point cloud. Both suggest that three clusters is a reasonable interpretation of the structure, but layout e more accurately reflects the separation between them, which is small.

@@

```{r}
#| fig-cap: "Assessing which of the 8 NLDR layouts on the PBMC3k data  (shown in @fig-NLDR-variety) is the better representation using RMSE for varying binwidth ($a_1$). Color used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-h). Layout f is universally poor. Layouts a, c, g, h that show large separations between clusters are universally suboptimal. Layout d with little separation performs well at tiny binwidth (where most points are in their own bin) and poorly as binwidth increases. The choice of best is between layouts b and e, that have small separations between oddly shaped clusters. Layout e is the best choice."
#| label: fig-pbmc-rmse
#| fig-pos: H
#| out-width: 100%

>>>>

In this data there are $2622$ single cells and $1000$ gene expressions (variables). Following the same pre-processing as @chen2024, different NLDR techniques were performed on the first nine principal components. @fig-NLDR-variety shows this data using a variety of methods, and different hyper-parameters. You can see that the result is wildly different depending on the choices. Layout a is a reproduction of the layout that was published in @chen2024. This layout suggests that the data has three very well separated clusters, each with an odd shape. The question is whether this accurately represents the cluster structure in the data, or whether they should have chosen b or c or d or e or f or g or h. This is what our new method can help with -- to decide which is the more accurate \gD{} representation of the cluster structure in the \pD{} data.

@@

To further assess the choices, we need to look at the model in the data space, by using a tour to show the wireframe model overlaid on the data in the $9\text{-}D$ space (@fig-model-pbmc-author-proj). Here we compare the published layout (a) versus what we argue is the best layout (e). The top row (a1, a2, a3) correspond to the published layout and the bottom row (e1, e2, e3) correspond to the optimal choice according to our procedure. The middle and right plots show two projections. The primary difference between the two models is that the model of layout e does not fill out to the extent of the data but concentrates in the center of each point cloud. Both suggest that three clusters is a reasonable interpretation of the structure, but layout e more accurately reflects the separation between them, which is small.

@@

```{r}
#| fig-cap: "Assessing which of the 8 NLDR layouts on the PBMC3k data  (shown in @fig-NLDR-variety) is the better representation using RMSE for varying binwidth ($a_1$). Color used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-h). Layout f is universally poor. Layouts a, c, g, h that show large separations between clusters are universally suboptimal. Layout d with little separation performs well at tiny binwidth (where most points are in their own bin) and poorly as binwidth increases. The choice of best is between layouts b and e, that have small separations between oddly shaped clusters. Layout e is chosen as the best."
#| label: fig-pbmc-rmse
#| fig-pos: H
#| out-width: 100%
